### 2014-11-02 application.tex
Add more text in MiniMax.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=&\[\]]
type route edge(node, node, int).
type linear shortest(node, int, list int).
type linear relax(node, int, list int).

&underline[priority @order asc].

shortest(A, +00, \[\]).
relax(@1, 0, [@1]).

relax(A, D1, P1), shortest(A, D2, P2),
D1 < D2
   -o shortest(A, D1, P1),
      {B, W | !edge(A, B, W) |
         relax(B, D1 + W, P1 ++ \[B\]),
         &underline[set-priority(B, float(D1 + W))]}.

relax(A, D1, P1), shortest(A, D2, P2),
D1 >= D2
   -o path(A, D2, P2).
\end{Verbatim}
  \caption{Shortest Path Program.}
  \label{code:shortest_path_program_coord}
\end{figure}
\normalsize

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{\includegraphics[width=4cm]{results/shortest-uspowergrid.png}}
      \subfloat[]{ \scriptsize\begin{tabular}{ | c | c | c | }
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 333K & 206K \\
         2 & 300K & 210K \\
         4 & 316K & 208K \\
         8 & 328K & 211K \\
         16 & 343K & 212K \\
         \hline  
         \end{tabular}
         \normalsize
      }
   \end{center}
   \caption{Single Source Shortest Path program code.}
   \label{results:sssp_uspowergrid}
\end{figure}

\subsection{Graph Search}

@@

\begin{figure}[h!]
   \begin{center}
      \includegraphics[width=5cm]{figures/minimax_tree}
   \end{center}
   \caption{Single Source Shortest Path program code.}
   \label{minimax}
\end{figure}

>>>>

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
type route edge(node, node, int).
type linear shortest(node, int, list int).
type linear relax(node, int, list int).

\underline{priority @order asc}.

shortest(A, +00, []).
relax(@1, 0, [@1]).

relax(A, D1, P1), shortest(A, D2, P2),
D1 < D2
   -o shortest(A, D1, P1),
      \{B, W | !edge(A, B, W) |
         relax(B, D1 + W, P1 ++ [B]),
         \underline{set-priority(B, float(D1 + W))}\}.

relax(A, D1, P1), shortest(A, D2, P2),
D1 >= D2
   -o shortest(A, D2, P2).
\end{Verbatim}
  \caption{Shortest Path Program.}
  \label{code:shortest_path_program_coord}
\end{figure}
\normalsize

The most interesting property of the SSSP program presented in
Fig.~\ref{code:shortest_path_program_coord} is that it remains provably correct,
although it applies rules using smarter ordering. The derivation of
\texttt{set-priority} does not change the behavior of the logical rules and the
code remains declarative.

Figure~\ref{results:sssp_uspowergrid} shows experimental results for a variant of
the SSSP program that computes the SSSP for many of the nodes of the graph.
There are some situations where unecessary facts are propagated
because although the shortest distance is selected, sub-optimal distances may be
propagated because many SSSP distances are computed at the same time.
However, we see a reduction of over 50\% in the number of
derived facts when using the coordinated version \textbf{C} over the regular
version \textbf{R}. The results also show that the cordinated
version using 16 threads is 1.3 times faster than the regular version using 16
threads. In turn, this makes the coordinated version 20 times faster
than the regular version using 1 thread.

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{ \begin{tabular}[b]{ | c | c | c | }
         \hline                       
         \textbf{\# T} & \textbf{R} & \textbf{C} \\ \hline \hline
         1 & 333K & 206K \\
         2 & 300K & 210K \\
         4 & 316K & 208K \\
         8 & 328K & 211K \\
         16 & 343K & 212K \\
         \hline  
         \end{tabular}
         \normalsize
      }
      \subfloat[]{\includegraphics[width=5cm]{results/shortest-uspowergrid.png}}
   \end{center}
   \caption{Experimental results for the SSSP program using US powergrid network.
   On the left we show the number of facts derived for the regular \textbf{R}
   and coordinated version \textbf{C} using a variable number of threads
   \textbf{\# T}. On the right, we have the scalability of the regular version
   and coordinated version. The speedup values are computed using the execution
   time of the regular version for 1 thread.}
   \label{results:sssp_uspowergrid}
\end{figure}

@@

The MiniMax algorithm is a decision rule algorithm for minimizing the
possible loss for a worst case (maximum loss) scenario in a zero sum
game for 2 (or more) players that play in
turns~\cite{Edwards54}.

The algorithm proceeds by building a game tree, where each tree node represents a
game state and the children represent the possible game moves that can
be made by either player 1 or player 2. The
tree levels are built in a way that each player plays in turns and the
first level represents the moves made by player 1.  An
evaluation function \texttt{f(State)} is used to compute the score of
the board for each leaf of the tree. A node is a leaf when it can no longer be
expanded since the state represents a possible end of the game.  Finally, the
algorithm recursively minimizes or maximizes the scores of each node.
To select the best move for player 1, the
algorithm picks the move maximized at the root node.

In LM, the program starts with a root node (with the initial game state)
that is expanded with the available moves at each level. The graph of the
program is dynamic since nodes are created and eventually deleted once they
have nore more facts or no more references to them. The latter happens when the
leaves scores are computed or when a node fully minimizes or maximizes the
children scores. When the program ends, only the root node is have facts in its
database.

The LM code in Fig.~\ref{minimax:check-end} implements
the initial expansion of the tree. The first three rules (lines 1-10) deal
with the cases where no children nodes are created and the last three rules
(12-29) deal with cases where new nodes may be created.
Thes rule in lines 12-25 generate new nodes using the
\texttt{exists} language construct, which creates a new node address
\texttt{B}, where a new \texttt{play} fact is created with the new
board. A \texttt{parent(B,A)} fact is derived to allow node \texttt{B} to
communicate with its parent.

Since our implementation
uses a FIFO ordering for nodes, meaning that when children are expanded, we
first expand each children and then the children of the children. The default
ordering of the system works well enough for most practical purposes. We note
however, that this is not optimal for this program since the complete tree needs
to be expanded before computing the scores at the leaves. This introduces memory
issues since, in the worst case, the memory needs of the program are
$\mathcal{O}(n)$, where $n$ is the number of nodes in the tree.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\{\}]
expand(A, Board, [], 0, P, \underline{Depth})
  -o leaf(A, Board).

expand(A, Board, [], N, P, \underline{Depth}),
N > 0
  -o maximize(A, N, -00, 0).

expand(A, Board, [], N, P, \underline{Depth}),
N > 0, NextPlayer <> RootPlayer
  -o minimize(A, N, +00, 0).

expand(A, Board, [0 | Xs], N, P, Depth),
Depth >= 5
  -o exists B. (\underline{set-static(B)},
       \underline{set-default-priority(B, float(Depth + 1))},
       play(B, Board ++ [P | Xs], next(P), \underline{Depth + 1}),
       expand(A, Board ++ [0], Xs, N + 1, P, \underline{Depth}),
       parent(B, A)).

expand(A, Board, [0 | Xs], N, P, Depth),
Depth < 5
  -o exists B. (\underline{set-default-priority(B, float(Depth + 1))},
       play(B, Board ++ [P | Xs], next(P), \underline{Depth + 1}),
       expand(A, Board ++ [0], Xs, N + 1, P, \underline{Depth}),
       parent(B, A)).

expand(A, Board, [C | Xs], N, P, \underline{Depth})
C <> 0
  -o expand(A, Board ++ [C], Xs, N, P, \underline{Depth}).
\end{Verbatim}
\caption{MiniMax: checking if the game has ended and expanding the tree}
\label{minimax:check-end}
\end{figure}
\normalsize

With coordination, we set the priority nodes to be the same as the
depth so that the tree is expanded in a depth-first fashion (lines 15 and 22).
The memory complexity of the program then becomes $\mathcal{O}(d t)$, where $d$
is the depth of the tree and $t$ is the number of threads. In
Fig.~\ref{fig:minimax}, we have two threads (squares) and the first one has
expanded the tree on the left side. When a second thread comes up, it steals the
first unexpanded node and expands the tree from that position. Since threads
prioritize deeper nodes, the scores of the first leaves are immediatelly
computed and then sent to the parent node. At this point, the leaves are deleted
and reused for other nodes in the tree, resulting in minimal memory usage.

\begin{figure}[h!]
   \begin{center}
      \includegraphics[width=4.5cm]{figures/minimax_tree}
   \end{center}
   \caption{Expanding the MiniMax tree using coordination. By prioritizing
      deeper nodes, threads are forced to expand the tree using a depth-first
      approach, which is superior since there is no need to expand the whole
      tree before computing the node scores.}
   \label{fig:minimax}
\end{figure}

We also take advantage of memory locality by using \texttt{set-static} (line
14), so that nodes after a certain level are not stolen by other threads. While
this is not critical for performance in shared memory systems where node
stealing is fairly efficient, we expect that such coordination to be critical in
distributed systems.
